
Set my working directory to this project folder because for some reason it wasn't before
```{r setup, include=FALSE}
setwd("/projectnb/bf528/students/jgsherry/project-2-JackSherry6")
```

Load in some libraries
```{r}
library(DESeq2)
library(tidyverse)
library(dplyr)
```

Read in gene expression data for all samples from csv file
```{r}
comb_matrix <- read_csv("results/combined_matrix.csv")
comb_matrix
```


Filter out genes with low expression levels
```{r}
keep <- rowSums(comb_matrix >= 10) >= 2
#keep <- comb_matrix[rowSums(comb_matrix == 0) != ncol(comb_matrix), ]
counts_filtered <- comb_matrix[keep, ]
#reorder gene columns below
counts_filtered <- counts_filtered[, c("gene","control_rep1","control_rep2","control_rep3","exp_rep1","exp_rep2","exp_rep3")]
counts_filtered <- counts_filtered %>% column_to_rownames(var = "gene") %>%  as.matrix() 
nrow(counts_filtered)
head(counts_filtered)
```




Create a dataframe of column data to use in our DESeq function
```{r}
colData <- data.frame(
  sample = c("control_rep1","control_rep2","control_rep3","exp_rep1","exp_rep2","exp_rep3"),
  group = c("control","control","control","exp","exp","exp")
)
colData
```


Run DESEQ on our filtered count data by group (control/exp)
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_filtered, 
                                colData = colData, 
                                design = ~ group)
```

Finish running DESeq
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast=c("group","control","exp"))
```

Format the DESeq data in a tible dataframe
```{r}
resOrdered <- res[order(res$padj),]
deseq2_res <- as_tibble(resOrdered, rownames='gene')
deseq2_res
```


read gene_names csv file in as a df but change the name of the column to match my DESeq df
```{r}
ids <- read_tsv("results/gene_names.csv")
names(ids)[1] <- "gene"
ids 
```


Add the gene names to the DESeq dataframe at the second column
```{r}
joint_df <- left_join(deseq2_res, ids, by="gene")
joint_df <- joint_df %>% relocate(gene_name, .after = gene)
joint_df
```


Order the df based on the adjusted p-value and filter out adjusted p-values above 0.05
```{r}
df_padj <- joint_df %>% arrange(padj)
filter_df <- df_padj %>% filter(padj < 0.05)
filter_df
```
After filtering for genes with a padj below 0.05, I am left with 133 genes for my DAVID and ENRICHR analyses.


```{r}
test_df <- filter_df %>% drop_na()
test_df
```

Running ENRICHR
```{r}
library(enrichR)
dbs_to_use <- c("GO_Biological_Process_2023", "KEGG_2021_Human")
gene_list <- filter_df$gene_name
enrich_results <- enrichr(gene_list, dbs_to_use)
head(enrich_results)
```

Saving GO and KEGG files 
```{r}
write.csv(enrich_results[["GO_Biological_Process_2023"]],
          "GO_BP_enrichr_results.csv", row.names = FALSE)
write.csv(enrich_results[["KEGG_2021_Human"]],
          "KEGG_enrichr_results.csv", row.names = FALSE)
```

create a top hits dataframe from GO and visualize (only shows top 20)
```{r}
library(ggplot2)
go_df <- read.csv("GO_BP_enrichr_results.csv")
top_hits <- go_df[order(go_df$Adjusted.P.value), ][1:20, ]
ggplot(top_hits, aes(x = reorder(Term, -Combined.Score), y = Combined.Score)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "GO Term", y = "Combined Score", title = "Top Enriched GO Terms") +
  theme_minimal(base_size = 12)
```

Create a top hits dataframe from KEGG and visualize (only shows top 20)
```{r}
enrich_df <- read.csv("KEGG_enrichr_results.csv")
top_hits <- enrich_df[order(enrich_df$Adjusted.P.value), ][1:20, ]
ggplot(top_hits, aes(x = reorder(Term, -Combined.Score), y = Combined.Score)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "KEGG term", y = "Combined Score", title = "Top Enriched KEGG Terms") +
  theme_minimal(base_size = 12)
```

RNAseq Quality Control using Variance Stabilizing Transformation (VST)
```{r}
vsd <- vst(dds, blind = TRUE)
norm_mat <- assay(vsd)
```

Get summary statistics table for pca
```{r}
pca <- prcomp(t(norm_mat))  # transpose so samples are rows
summary(pca)
```

Sample-to-sample distance matrix:
```{r}
library(pheatmap)
library(RColorBrewer)

sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         main = "Sample-to-sample distances")
```

PCA plot with labels
```{r}
#plotPCA(vsd, intgroup = "group")
library(ggplot2)
p <- plotPCA(vsd, intgroup = "group", returnData = TRUE)  # get the data
percentVar <- round(100 * attr(p, "percentVar"))           # get variance for axis labels

# Make PCA plot with labels
ggplot(p, aes(PC1, PC2, color = group, label = name)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()
``` 


Starting FGSEA analysis
```{r}
library(fgsea)
```

Create a df that hasn't been filtered by padj and is ordered by log2FoldChange descending
Notes: removed dropNA because they should all be gone
```{r}
deseq2_res_ordered <- joint_df %>% 
  drop_na() %>%
  distinct(gene_name, .keep_all = TRUE) %>%
  arrange(desc(log2FoldChange))

fgsea_list <- setNames(deseq2_res_ordered$log2FoldChange, deseq2_res_ordered$gene_name)
head(fgsea_list, 10)
```


I scped the C2 canonical pathways file from GSEA website to this directory.
```{r}
pathways <- gmtPathways("C2_canonical_pathways.gmt")
head(pathways)
```

```{r}
fgseaRes <- fgsea(pathways, fgsea_list, minSize=15, maxSize=500)
head(fgseaRes[order(pval), ])
```


Get only adjusted p-values below 0.05
```{r}
sig_fgsea <- fgseaRes %>% 
  as_tibble() %>% 
  filter(padj < 0.05) %>%  
  arrange(padj)

sig_fgsea
```

```{r}
# make a variable containing only the pathways we want to plot based on our significance results
allPathways <- sig_fgsea$pathway[order(sig_fgsea$NES, decreasing = TRUE)]

# Plot all pathways ordered by NES
plotGseaTable(pathways[allPathways], fgsea_list, sig_fgsea, gseaParam = 0.5)
```

Figure 3F:
```{r}
alt_fgsea <- fgseaRes %>% 
  as_tibble() %>% 
  filter(padj < 0.25) %>%  
  arrange(padj)

ggplot(alt_fgsea, aes(x = NES, y = reorder(pathway, NES))) +
  geom_point(aes(size = -log10(padj), color = NES)) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red") +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathway", size = "-log10(padj)", color = "NES",
      title = "Top Enriched Pathways (fgsea)") +
  theme_bw()
```
Was unable to get the colors swapped but it should convey the same information.

Volcano plot:
```{r}
# setting these has been the only way I can get the labels to fit correctly (probably not the most efficient way though)
padj_threshold <- 0.05
log2FC_threshold <- 1

# Add a color column for left/right coloring
filter_df$Color <- ifelse(
  filter_df$padj < padj_threshold & filter_df$log2FoldChange < -log2FC_threshold, "blue",
  ifelse(filter_df$padj < padj_threshold & filter_df$log2FoldChange > log2FC_threshold, "pink", "grey")
)

# Top 
top_up <- filter_df[filter_df$Color == "pink", ]
top_up <- top_up[order(top_up$padj), ][1:min(10, nrow(top_up)), ]

top_down <- filter_df[filter_df$Color == "blue", ]
top_down <- top_down[order(top_down$padj), ][1:min(10, nrow(top_down)), ]

top_genes <- rbind(top_up, top_down)

ggplot(filter_df, aes(x = log2FoldChange, y = -log10(padj), color = Color)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_text(
    data = top_genes,
    aes(label = gene_name),
    vjust = -0.5, # vertical adjustment
    hjust = 0.5,  # horizontal adjustment
    size = 3
  ) +
  scale_color_manual(values = c("blue" = "blue", "pink" = "pink", "grey" = "grey")) +
  theme_minimal() +
  xlab("Expression Difference (log2 Fold Change)") +
  ylab("Significance (-log10 Adjusted p-value)") +
  ggtitle("Volcano Plot of Differential Expression") +
  theme(legend.title = element_blank())
```
Compare how many significant genes are up- and down-regulated in their findings and yours (using their significance threshold). 
Ensure you list how many you find vs. how many they report.

